---
title: "Mini Data Analysis Milestone 2"
output: github_document
---


*To complete this milestone, you can either edit [this `.rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-2.Rmd) directly. Fill in the sections that are commented out with `<!--- start your work here--->`. When you are done, make sure to knit to an `.md` file by changing the output in the YAML header to `github_document`, before submitting a tagged release on canvas.*

# Welcome back to your mini data analysis project!

This time, we will explore more in depth the concept of *tidy data*, and hopefully investigate further into your research questions that you defined in milestone 1.

**NOTE**: The main purpose of the mini data analysis is to integrate what you learn in class in an analysis. Although each milestone provides a framework for you to conduct your analysis, it's possible that you might find the instructions too rigid for your data set. If this is the case, you may deviate from the instructions -- just make sure you're demonstrating a wide range of tools and techniques taught in this class.

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
```

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with manipulating and summarizing your data in tibbles using `dplyr` and `tidyr`, with a research question in mind.
+ Understand what *tidy* data is, and how to create it. In milestone 3, we will explore when this might be useful.
+ Generate a reproducible and clear report using R Markdown.
+ Gain a greater understanding of how to use R to answer research questions about your data.

**Things to keep in mind**

+ Remember to document your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. Create your analysis as if someone else will be reading it! **There will be 2.5 points reserved for reproducibility, readability, and repo organization.**

+ Before working on each task, you should always keep in mind the specific **research question** that you're trying to answer.

# Task 1: Process and summarize your data (15 points)

From milestone 1, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions. 

### 1.1 (2.5 points) 

First, write out the 4 research questions you defined in milestone 1 were. This will guide your work through milestone 2:

<!-------------------------- Start your work below ---------------------------->
1. *Which parameters are highly correlated to the diagnosis types?*
2. *Whether the relationship between cancer sample parameters and diagnosis type is consistent for mean, SE, and worst*
3. *What would be the relationship between means and worsts?*
4. *What would be the relationship between SEs and worsts?*
<!----------------------------------------------------------------------------->

### 1.2 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question. 

Ensure that the output of each operation is printed! 

**Summarizing:**

1. Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read. 
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
1. *Which parameters are highly correlated to the diagnosis types?*
#Regarding this question, we can actually compare the summary statistics of one variable across the diagnosis groups. If that parameter is highly correlated to the diagnosis types, we would see large differences between summary statistics across diagnosis type#

```{r, message = FALSE}

#compute the mean of radius_mean across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(mean_radius_mean = mean(radius_mean, na.rm = TRUE))

#compute the range of radius_mean across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(range_radius_mean = range(radius_mean, na.rm = TRUE))

#compute the median of radius_mean across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(median_radius_mean = median(radius_mean, na.rm = TRUE))

#compute the SD of radius_mean across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(sd_radius_mean = sd(radius_mean, na.rm = TRUE))
```
#it seems that the summary statistics differ significantly for radius_mean, and patients with diagnosis type M tends to have larger mean, larger range, larger median and larger SD values compared to patients with diagnosis type B. 

#with density graph, we can also observe how the radius_mean differ for diagnosis type
#I use alpha = 0.3 because it allows me visualize whether there is overlap for radius_mean for both diagnosis type

```{r, message = FALSE}
ggplot(cancer_sample, aes(x = radius_mean)) + 
    geom_density(aes(fill = diagnosis),alpha=0.3)#density plot for radius_mean group by diagnosis type
```
#even there is overlap of density between radius_mean, but it is clear that patients diagnosed with M have relatively higher radius_mean compared to those patients diagnosed with B. This indicates that radius_mean is related to diagnosis type and could use as a parameter to estimate diagnosis type. 

2. *Whether the relationship between cancer sample parameters and diagnosis type is consistent for mean, SE, and worst*

#we can keep working on the parameter 'radius' and see whether its SD and worst values also vary significantly across diagnosis type#
```{r, message = FALSE}
#let's take a look at the summary statistics of radius_se#
#compute the mean of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(mean_radius_se = mean(radius_se, na.rm = TRUE))

#compute the range of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(range_radius_se = range(radius_se, na.rm = TRUE))

#compute the median of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(median_radius_se = median(radius_se, na.rm = TRUE))

#compute the SD of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(sd_radius_se = sd(radius_se, na.rm = TRUE))

#let's take a look at the summary statistics of radius_worst#
#compute the mean of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(mean_radius_worst = mean(radius_worst, na.rm = TRUE))

#compute the range of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(range_radius_worst = range(radius_worst, na.rm = TRUE))

#compute the median of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(median_radius_worst = median(radius_worst, na.rm = TRUE))

#compute the SD of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(sd_radius_worst = sd(radius_worst, na.rm = TRUE))

```
#it seems that the summary statistics differ significantly for both radius_se and radius_worst, and patients with diagnosis type M tends to have larger mean, larger range, larger median and larger SD values compared to patients with diagnosis type B. The relationship between radius and diagnosis type is consistent for mean, SE, and worst. 

#with density graph, we can also observe how the radius_se and radius_worst differ for diagnosis type
#I use alpha = 0.3 because it allows me visualize whether there is overlap for radius_Se and radius_worst values for both diagnosis type
```{r, message = FALSE}
ggplot(cancer_sample, aes(x = radius_se)) + 
    geom_density(aes(fill = diagnosis),alpha=0.3)#density plot for radius_se

ggplot(cancer_sample, aes(x = radius_worst)) + 
    geom_density(aes(fill = diagnosis),alpha=0.3)#density plot for radius_worst
```
#even there is overlap of density between both radius_se and radius_worst, but it is clear that patients diagnosed with M have relatively higher radius_worst and radius_se compared to those patients diagnosed with B. This further verify our conclusion that The relationship between radius and diagnosis type is consistent for mean, SE, and worst. 


3. *What would be the relationship between means and worsts?*
#since we have computed the summary statistics for radius_mean, we can compute the summary statistics for radius_mean
```{r, message = FALSE}

#compute the mean of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(mean_radius_worst = mean(radius_worst, na.rm = TRUE))

#compute the range of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(range_radius_worst = range(radius_worst, na.rm = TRUE))

#compute the median of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(median_radius_worst = median(radius_worst, na.rm = TRUE))

#compute the SD of radius_worst across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(sd_radius_worst = sd(radius_worst, na.rm = TRUE))
```
#While comparing to the summary statitics for radius_mean, we found that radius_worst has slightly higher mean, median and standard deviation values compared to radius_mean, and the range for radius_worst is also slightly larger in type M compared to the range for radius_mean while the range for radius_worst is almost the same with the range for radius mean in type B. 


#we can visualize the relationship between radius_mean and radius_worst with the scatterplot
```{r, message = FALSE}
ggplot(cancer_sample, aes(x = radius_mean, y = radius_worst)) +
       geom_point(aes(fill = factor(diagnosis)), shape = 21)#plot scatterplot between radius_mean and radius_worst
```
#I see a strong linear relationship (positive association) between radius_worst and radius_mean for type B, while slighter weaker linear relationship (positive association)  between radius_worst and radius_mean for type M. 

4. *What would be the relationship between SEs and worsts?*
#since we have computed the summary statistics for radius_worst, we can compute the summary statistics for radius_se
```{r, message = FALSE}

#let's take a look at the summary statistics of radius_se#
#compute the mean of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(mean_radius_se = mean(radius_se, na.rm = TRUE))

#compute the range of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(range_radius_se = range(radius_se, na.rm = TRUE))

#compute the median of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(median_radius_se = median(radius_se, na.rm = TRUE))

#compute the SD of radius_se across diagnosis type#
cancer_sample %>%
  group_by(diagnosis) %>%
  summarise(sd_radius_se = sd(radius_se, na.rm = TRUE))
```
#we know that patients diagnosed with M would have higher radius_se mean, median, standard deviation and range compared to patients diagnised with B. Since radius_se and radius_worst use different units, we could not compare them directly, but we can assume that linear relationship (positive association) could be observed for radius_se and radius_worst 

#we can visualize the relationship between radius_se and radius_worst with the scatterplot
```{r, message = FALSE}
ggplot(cancer_sample, aes(x = radius_se, y = radius_worst)) +
       geom_point(aes(fill = factor(diagnosis)), shape = 21)#plot scatterplot between radius_worst and radius_se
```
#However, according to the plot, there doesn't appear to be a strong positive association between radius_worst and radius se. We can even say that radius_se and radius_worst are independent of each other. 
<!----------------------------------------------------------------------------->

### 1.3 (2.5 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!------------------------- Write your answer here ---------------------------->
#Besides radius, there are also other parameters including texture, smoothness, compactness, concavity, concave_points, symmetry, and fractal_dimension. The patterns we observed for radius might not be applied to other parameters. I would need to explore all of the parameters to draw conclusion for each question. The unclear part related to my research is that whether the association assesed for each parameters happen randomly or it has any statistical significance. I believe this question would require further statistical test. My research questions are really general, and I would need to assess all the parameters to answer this questions. But currenly, I would still keep them because these questions would guide me through the exploration of those parameters. It is interesting that I thought there might be positive association between radius_se and radius_worst, but when I plotted the graph, it appears to be no association. It indicates that scatterplot is really helpful for indicating relationship of two varaibles. 

<!----------------------------------------------------------------------------->

# Task 2: Tidy your data (12.5 points)

In this task, we will do several exercises to reshape our data. The goal here is to understand how to do this reshaping with the `tidyr` package.

A reminder of the definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 

### 2.1 (2.5 points)

Based on the definition above, can you identify if your data is tidy or untidy? Go through all your columns, or if you have >8 variables, just pick 8, and explain whether the data is untidy or tidy.

<!--------------------------- Start your work below --------------------------->
The data is untidy. It violates the second principle that "each column is a **variable**". Column headers radius_mean, radius_se, and radius_wort are values, not variable names.These values could be put under the variable named radius. Additionally, concavity_mean, concavity_se and concavity_worst are also valies, not variable names, and these should be put under the variable named concavity. symmetry_mean, symmetry_se, and symmetry_worst are also values, they should also be put under the variable named symmetry. 

<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Now, if your data is tidy, untidy it! Then, tidy it back to it's original state.

If your data is untidy, then tidy it! Then, untidy it back to it's original state.

Be sure to explain your reasoning for this task. Show us the "before" and "after".

<!--------------------------- Start your work below --------------------------->
```{r, message = FALSE}
#for this task, we will try to convert the untidy data to tidy data#
#to achieve this goal, we will expect to see measures like mean, se, and worst stay under one column, and the values goes under the column with parameters name.

#let's view the "before" version where we have values like radius_mean, radius_se, radius_worst shown as columns
glimpse(cancer_sample)

#the "after" version will have measures like mean, se, and worst stay under one column, and the values goes under the column with parameters name.
tidy_cancer_sample<-cancer_sample %>% 
  rename(concave.points_mean = concave_points_mean, concave.points_se=concave_points_se,
         concave.points_worst=concave_points_worst, fractal.dimension_se=fractal_dimension_se, fractal.dimension_mean=fractal_dimension_mean, fractal.dimension_worst=fractal_dimension_worst)%>% #rename some variables to format the column names
  pivot_longer(cols      = c(-ID, -diagnosis), 
               names_to  = c(".value", "measure"),
               names_sep = "_")
```
<!----------------------------------------------------------------------------->

### 2.3 (5 points)

Now, you should be more familiar with your data, and also have made progress in answering your research questions. Based on your interest, and your analyses, pick 2 of the 4 research questions to continue your analysis in milestone 3, and explain your decision. 

Try to choose a version of your data that you think will be appropriate to answer these 2 questions in milestone 3. Use between 4 and 8 functions that we've covered so far (i.e. by filtering, cleaning, tidy'ing, dropping irrelvant columns, etc.). 

<!--------------------------- Start your work below --------------------------->

```{r, message = FALSE}
#The questions I am attempting to answer would be 
#1. *Which parameters are highly correlated to the diagnosis types?*
#2. *Whether the relationship between cancer sample parameters and diagnosis type is consistent for mean, SE, and worst*


#To answer this question, I would first tidy my dataset based on the tidy_cancer_sample, and then calculate the mean summary statistics based on diagnosis type and parameter, and see how the measures for each parameter differ for diagnosis types

mean<-tidy_cancer_sample%>%  
  filter(measure == "mean")%>%
  pivot_longer(cols = c(-ID, -diagnosis, -measure), 
               names_to  = "parameter", 
               values_to = "mean_value")%>%
  group_by(parameter, diagnosis)%>%summarise(mean_mean = mean(mean_value, na.rm = TRUE))

#Based on the table, it seems that patients diagnosed with M have higher average values in radius_mean, area_mean, compactness_mean, concave.points_mean, concavity_mean, perimeter_mean, 	symmetry_mean, and texture_mean compared to patients diagnosed with B, while their fractal.dimension_mean and smoothness_mean almost stays the same.


se<-tidy_cancer_sample%>%  
  filter(measure == "se")%>%
  pivot_longer(cols = c(-ID, -diagnosis, -measure), 
               names_to  = "parameter", 
               values_to = "se_value")%>%
  group_by(parameter, diagnosis)%>%summarise(se_mean = mean(se_value, na.rm = TRUE))

#Based on the table, it seems that patients diagnosed with M have higher average values in radius_se, area_se, compactness_se, concave.points_se, concavity_se, perimeter_se compared to patients diagnosed with B, while their fractal.dimension_mean, smoothness_mean, symmetry_se, and texture_se almost stays the same.

worst<-tidy_cancer_sample%>%  
  filter(measure == "worst")%>%
  pivot_longer(cols = c(-ID, -diagnosis, -measure), 
               names_to  = "parameter", 
               values_to = "worst_value")%>%
  group_by(parameter, diagnosis)%>%summarise(worst_mean = mean(worst_value, na.rm = TRUE))
  
#Based on the table, it seems that patients diagnosed with M have higher average values in radius_worst, area_worst, compactness_worst, concave.points_worst, concavity_worst, perimeter_worst fractal.dimension_worst, smoothness_worst, symmetry_worst, and texture_worst compared to patients diagnosed with B. 


#Conclusion
#Q1: According to the results, we can conclude that parameters like radius, area, compactness, concave.points, concavity and perimeter could be are highly correlated to diagnosis type, we could see their mean, worst and se measure vary significantly for different diagnosis types. 

#Q2:  For parameters like radius, area, compactness, concave.points, concavity and perimeter, we can see that patients diagnosed with M have higher average in mean, worst and se compared to patients diagnosed with B, indicating consistent relationship. However for parameters like fractal.dimension, smoothness, symmetry, and texture, we might see higher avarage value for people diagnosed with "M" in the worst group but when we look at the se group, the average values are almost the same for both diagnosis types, indicating inconsistent relationship. 

```
<!----------------------------------------------------------------------------->

*When you are done, knit an `md` file. This is what we will mark! Make sure to open it and check that everything has knitted correctly before submitting your tagged release.*

### Attribution

Thanks to Victor Yuan for mostly putting this together. 
